---
title: "CS109 Lab 1 Supplement: GAM FAQs"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---


There are two common problems that often occur when using the `gam`
package in R. This document shows you how to avoid them. First set up
an example used to demonstrate the problems:

```{r}
library(gam)
train.ind <- sample(1:nrow(cars), 25)
cars.train <- cars[train.ind, ]
cars.test <- cars[-train.ind, ]
```

Can't predict with new data
===========================

`predict.gam` does some complicated gymnastics behind the scenes, and
it sometimes gets tripped up. Specifically, passing arguments as
variable names in the `gam` formula argument results in a scoping
error. 

Here is what happens:

```{r, error = TRUE}
fit_gam <- function(spval) {
    gam(speed ~ s(dist, spar = spval),
        data = cars.train)
}

mp <- predict(fit_gam(spval = .2), newdata = cars.test)
```

The solution is to substitute the value of the variable you are
passing in. You can do that using (among other things), `substitute`
or `paste`:

```{r}
## pass spar value using substitute
fit_gam <- function(spval) {
    gam(as.formula(substitute(speed ~ s(dist, spar = spval), list(spval = spval))),
        data = cars.train)
}

mp <- predict(fit_gam(spval = .2), newdata = cars.test)
```

```{r}
## alternatvely, constuct the formula from a character vector
fit_gam <- function(spval) {
    gam(as.formula(paste("speed ~ s(dist, spar = ", spval, ")")),
        data = cars.train)
}

mp <- predict(fit_gam(spval = .2), newdata = cars.test)
```

`plot.gam` doesn't work
=======================

Problems plotting `gam` package models usually occurs because of
conflicts with the `mgcv` package. This conflict affects other things
as well, be we'll use `plot.gam` as the example.

Since many other packages use `mgcv` internally it is not always
obvious when you've triggered the conflict.

Usually `plot.gam` works just fine:

```{r}
plot(fit_gam(spval = .2))
```

Once the `mgcv` name space is loaded it breaks:

```{r, results = "hide", warning = FALSE}
library(ggplot2)
ggplot(cars, aes(x = dist, y = speed)) +
    geom_smooth(method = "gam") ## this loads the mgcv namespace
```

Now that the `mgcv` name space is loaded `plot.gam` (and probably
other stuff) is broken.

```{r}
any(grepl("mgcv", loadedNamespaces()))
```

```{r, error = TRUE}
plot(fit_gam(spval = .2))
```

The worst part is that the only way to recover is to restart your R
session. Unloading the name space should work, but it doesn't:

```{r}
unloadNamespace("mgcv")
any(grepl("mgcv", loadedNamespaces()))
```

```{r, error = TRUE}
plot(fit_gam(spval = .2))
```

If you are using the `gam` package and `any(grepl("mgcv",
loadedNamespaces()))` reports `TRUE`, you must restart your R session
and take steps to avoid loading `mgcv`. This includes making sure
`ggplot2` isn't loading it to make predictions with `geom_smooth`.

Since you probably want to use `ggplot` to plot prediction lines from
`gam` models you'll need a work-around to do that without loading
`mgcv`. The safest way is just to do the predictions yourself:


```{r, eval = FALSE, warning = FALSE}
library(ggplot2)
ggplot(transform(cars, pred = predict(gam(speed ~ s(dist, spar = .2), data = cars))),
       mapping = aes(x = dist, y = speed)) +
    geom_point() =
    geom_line(mapping = aes(y = pred))
```
