---
title: "CS109B Lab 3: Bayes Models"
output: 
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
---

Overview 
========

There are three "levels" so to speak at which you can perform a
Bayesian analysis in R. 

1. You can perform simple analyses using distribution, math, and
   simulation functions.
2. You can use packages providing interfaces to pre-coded standard
   models.
3. You can use R as an interface to a general purpose MCMC environment.

Level 1 is good for learning basic Bayesian principles and for
relatively simple analyses. Level 2 is good for fitting Bayesian
versions of many models you are probably familiar with in a
frequentest context. Finally, Level 3 is good for more complex
analyses for which the full power and flexibility of a MCMC approach
is needed.

In this lab session we'll look at Bayesian analyses at each of these
levels. But first...

Distribution functions in R
---------------------------

Bayesian data analysis requires specifying prior distributions for the
parameters in the model, then computing and analyzing posterior
distributions.

Base *R* (specifically the `stats` package) provides
        
          Density, cumulative distribution function, quantile function and
          random variate generation for many standard probability
          distributions

Refer to `?Distributions` for more information. Additional
distribution functions are available in various R packages.

You can read about the properties of various distributions, but if
you're a visual learner like me it can be useful to graph properties
of a distribution at varying parameter values to get an intuitive feel
for how the distribution behaves. For example, if we are not familiar
with the `beta` distribution:

```{r}
shapes <- round(2^(3:8)/10, digits = 2)
x <- seq(0, 1, by = 0.01)

## all combinations of x, alpha, and beta
beta_d <- expand.grid(x = x,
                      alpha = shapes,
                      beta = shapes)

## split the data by alpha and beta, calculate density, then re-assemble (rbind).
beta_d <- do.call(rbind,
                  lapply(split(beta_d, beta_d[c("alpha", "beta")]),
                         function(data) {
                             data.frame(data,
                                        density = dbeta(data$x,
                                                        unique(data$alpha),
                                                        unique(data$beta)))
                         }))

library(ggplot2)

ggplot(beta_d, aes(x = x, y = density)) +
    geom_line() +
    facet_grid(beta ~ alpha, labeller = label_both) +
    scale_y_continuous(limits = c(0, 15))
```

Simple analysis of proportions
==============================

*The lady tasting tea* is a classic in the statistic literature. The
basic premise is that a lady claims to be able to tell if milk or tea
was added to a cup first. Variations of the basic idea have appeared
in statistics texts ever since the original reported in R. Fisher's
*The Design of Experiments* (1935). In most descriptions of the
experiment the lady correctly identifies 8/8 cups of tea. What, in a
Bayesian framework, does that tell us?

We can work it out using basic distribution statistics.

```{r}
## Obseved data
obs <- c(1, 1, 1, 1, 1, 1, 1, 1)
obs.count <- table(factor(obs, levels = c("0", "1")))

## Possible values of the unknown parameter
x <- (0:100)/100

## priors, likelihood, and posterior expressed as beta distribution parameters
prior_alpha <- prior_beta <- 1 ## dbeta(x, 1, 1) == dunif(x)
likelihood_alpha <- obs.count["1"] + 1
likelihood_beta <- obs.count["0"] + 1
posterior_alpha <- obs.count["1"] + prior_alpha
posterior_beta <- obs.count["0"] + prior_beta

p_tea1 <- data.frame(x = x, # possible values of our unobserved parameter theta
                     prior = dbeta(x, prior_alpha, prior_beta), # uniformative prior (does this make sense?)
                     likelihood = dbeta(x, # x^8*(1-x)^0 scaled to match prior
                                        likelihood_alpha,
                                        likelihood_beta),
                     posterior = dbeta(x, # since the prior is uninformative the posterior distribution is the likelihood
                                       posterior_alpha,
                                       posterior_beta))

library(tidyr)

p_tea1 <- gather(p_tea1,
                key = "distribution",
                value = "density",
                -x)

library(ggplot2)

ggplot(p_tea1,
       mapping = aes(x = x, y = density, color = distribution)) +
    geom_line()
```

Does an uninformative prior make sense?
-----------------------------------

By specifying a uniform distribution we are saying that before seeing
the data $\theta$ has equal probability of being any value between
zero and 1. But that doesn't seem quite right. Even if she can not
tell the difference, the lady is likely to get some right. Perhaps
before seeing the data we expect the median outcome to be 4/8, and
that we are pretty confident about that. We can include this prior
information in the analysis using a beta distribution.

The LearnBayes package has a utility for choosing a beta distribution
that matches our prior beliefs.

```{r}
##  install.packages("LearnBayes")
library(LearnBayes)

prior_shape <- beta.select(list(p = .5, x = .5), # middle of the distribution is .5
                           list(p = .95, x = .7)) # 95th percentile is .7

## this is the beta distribution with mdeian = .5 and 95th percentile = .7
prior_shape
```

We can use that information to calculation the posterior distribution
again, incorporating this prior information.

```{r}
prior_alpha <- prior_shape[1]
prior_beta <- prior_shape[1]
likelihood_alpha <- obs.count["1"] + 1
likelihood_beta <- obs.count["0"] + 1
posterior_alpha <- obs.count["1"] + prior_alpha
posterior_beta <- obs.count["0"] + prior_beta

p_tea2 <- data.frame(x = x,
                     prior = dbeta(x, prior_alpha, prior_beta),
                     likelihood = dbeta(x, # x^8*(1-x)^0 scaled to match prior
                                        likelihood_alpha,
                                        likelihood_beta),
                     posterior = dbeta(x,
                                       posterior_alpha,
                                       posterior_beta))

p_tea2 <- gather(p_tea2,
                 key = "distribution",
                 value = "density",
                 -x)

ggplot(p_tea2,
       mapping = aes(x = x, y = density, color = distribution)) +
    geom_line() +
    scale_x_continuous(breaks = seq(0.0, 1.0, by = 0.1))
```

Analyzing the posterior distribution
------------------------------------

Once the posterior distribution has been obtained it can be analyzed to
obtain expected parameter values and other summary statistics. For the
beta distribution we can calculate these values analytically:

```{r}
theta_summary1 <- c(mode = (posterior_alpha - 1/3) /
                           (posterior_alpha + posterior_beta - 1),
                    median = (posterior_alpha - 1/3) /
                             (posterior_alpha + posterior_beta - (2/3)))

theta_summary1
```

 A useful and general alternative is to simulate a sample from the
 distribution and analyze the simulated values:

```{r}
theta <- rbeta(1000000,
               posterior_alpha,
               posterior_beta)

Mode <- function(x) {
    dx <- density(x)
    dx$x[which.max(dx$y)]
}

theta_summary2 <- c(mode = Mode(theta),
                    median = median(theta),
                    lq = quantile(theta, .025),
                    uq = quantile(theta, 0.975),
                    mean = mean(theta),
                    var = var(theta))
theta_summary2
 ```

Your Turn: Probability of a Purchase
===================================

Suppose that you run a website selling cat pajamas. You would like to
calculate the probability that your star salesperson closes a sale.
Over the course of a day you have 50 visitors, 5 of whom make a
purchase.

1. Choose and define an uninformative prior distribution for the
   parameter in your model.

2. Construct a likelihood function based on the observed 5/50
   successes.
   
3. Calculate the posterior distribution of the parameter.

4. Calculate the expected value of the parameter.

5. (Advanced) Plot the likelihood function and the prior and posterior
   distributions to see visually how we updated our beliefs about the
   probability based on the data.

6. (Advanced) Fit a logistic regression predicting sales (this is an
   intercept-only model) and compare the results to those obtained
   from the Bayesian analysis.


Classification example
======================

A classic classification example is the spam filter. I've made up some
data based on
<https://archive.ics.uci.edu/ml/machine-learning-databases/spambase/>
that we can use to illustrate.

```{r}
spam <- read.csv("http://tutorials.iq.harvard.edu/example_data/spam.csv")
str(spam)
```

Next write a function to calculate the probability of being in the
first class. (This function definition was copied from homework 3.)

```{r}
posterior_pA = function(alpha, yA = NULL, yB = NULL, y_til = NULL){
    ## number of features
    K = length(yA)
    ## total word counts
    n = sum(y_til)
    nA = sum(yA)
    nB = sum(yB)
    ## posterior predictive distribution of being class A
    A1 = lfactorial(n) + lfactorial(nA) - lfactorial(n + nA)
    A2 = sum(lfactorial(y_til + yA)) - sum(lfactorial(y_til)) - sum(lfactorial(yA))
    A3 = lfactorial(n + nA) + lgamma(K*alpha) - lgamma(n + nA + K*alpha)
    A4 = sum(lgamma(y_til + yA + alpha) - lfactorial(y_til + yA) - lgamma(alpha))
    A5 = lfactorial(nB) + lgamma(K*alpha) - lgamma(nB + K*alpha)
    A6 = sum(lgamma(yB + alpha) - lfactorial(yB) - lgamma(alpha))
    R_A = exp(A1 + A2 + A3 + A4 + A5 + A6)
    ## posterior predictive distribution of being class B
    B1 = lfactorial(n) + lfactorial(nB) - lfactorial(n + nB)
    B2 = sum(lfactorial(y_til + yB)) - sum(lfactorial(y_til)) - sum(lfactorial(yB))
    B3 = lfactorial(n + nB) + lgamma(K*alpha) - lgamma(n + nB + K*alpha)
    B4 = sum(lgamma(y_til + yB + alpha) - lfactorial(y_til + yB) - lgamma(alpha))
    B5 = lfactorial(nA) + lgamma(K*alpha) - lgamma(nA + K*alpha)
    B6 = sum(lgamma(yA + alpha) - lfactorial(yA) - lgamma(alpha))
    R_B = exp(B1 + B2 + B3 + B4 + B5 + B6)
    ## probability of being class A
    pA = R_A/(R_A + R_B)
    return(pA)
}
```

`posterior_pA` above generates probabilities for a single case. It
will be more useful if we wrap it so you can calculate probabilities
for a data.frame of cases.

```{r}
## posterior_pA above generates probabilities for a single case.
## here we wrap it so you can calculate probabilities for a data.frame of cases.
class_probability = function(data, class, alpha) {
    y <- lapply(split(data, class),
                function(x) {
                    apply(x, 2, sum)
                })
    yA <- y[[1]]
    yB <- y[[2]]
    pA <- apply(data, 1, function(x) {
        posterior_pA(alpha = alpha,
                     yA = yA,
                     yB = yB,
                     y_til = x)
    })
    pA
}

```

Finally we can calculate the probability that each message is spam.

```{r}
pa1 <- class_probability(spam[-1], spam[1], alpha = 1)

## confusion matrix, as proportions.
confmat <- function(actual, predicted) {
    prop.table(table(actual, predicted),
               margin = 2)
}

confmat(spam$spam,
        factor(pa1 < .5,
               labels = c("email", "spam")))
```

As a quick point of comparison, here is the accuracy of a naive Bayes
classifier.


```{r}
## install.packages("e1071")
library(e1071)

confmat(spam$spam,
        predict(naiveBayes(spam ~ ., data = spam), newdata = spam))

```

Your Turn: Predict Party from Voting Pattern
============================================

In this example you will predict party membership from voting
patterns. The data is available at 

<https://archive.ics.uci.edu/ml/machine-learning-databases/voting-records/house-votes-84.data>

and the documentation is at

<https://archive.ics.uci.edu/ml/machine-learning-databases/voting-records/house-votes-84.names>

1. Read the data into R and inspect them.

2. Convert `y` and `n` values to `1` and `0`.

3. Remove missing values. (Not a good idea in general!)

4. Use the `posterior_pA` function provided above to predict party
   membership from voting patterns.
   
5. Generate a confusion matrix to check the accuracy of your
   prediction.

6. (Advanced) Split the data into a train and test set. Predict party
   in the test based on voting patterns in the train.


Specific Bayes Models using MCMC
================================

For more complex models we will usually turn to MCMC as a means of
sampling from the posterior distribution. There are several packages
that provide Bayesian versions of common statistical models using
MCMC, including the `MCMCpack` and `rstanarm` packages.

Fitting MCMC models in R
------------------------

Let's look again at the problem of classifying messages as spam vs
email, this time using a Bayesian version of logistic regression.


```{r}
## install.packages("MCMCpack")
## install.packages(c("rstan", "rstanarm"))

library(rstanarm) # based on the stan probabilistic programming language
library(MCMCpack) # based on the scythe statistical library

# these approaches (esecially stan) reguire a lot of computation. Use multiple cores if you have them.
options(mc.cores = parallel::detectCores() - 1)
```

Let's first set up a training a test set of messages
```{r}
spam <- read.csv("http://tutorials.iq.harvard.edu/example_data/spam.csv")

intrain <- sample(1:nrow(spam), 300)
train <- spam[intrain, ]
test <- spam[-intrain, ]

```

Next we'll compare the accuracy of:

-   Logistic regression (`glm`)
-   Bayesian logistic regression (`stan_glm`)
-   Bayesian logistic regression ('MCMClogit`)


```{r}
fit0 <- glm(spam ~ ., data = train,
            family = binomial(link = "logit"))

confmat(test$spam,
        factor(predict(fit0,
                       newdata = test,
                       type = "response") > .5,
               labels = c("p_email", "p_spam")))

prop.table(table(factor(predict(fit0,
                                newdata = test,
                                type = "response") > .5,
                        labels = c("p_email", "p_spam")),
                 test$spam), margin = 1)
```

```{r}
fit1 <- stan_glm(spam ~ ., data = test, 
                 family = binomial(link = "logit"))

## we need to do a little work to get predictions.
pd1 <- posterior_predict(fit1,
                        newdata = test,
                        draws = 2000)

pd1 <- apply(pd1, 2, mean)

confmat(test$spam,
        factor(pd1 > .5,
               labels = c("email", "spam")))
```

```{r}
fit2 <- MCMClogit(spam ~ ., data = transform(train, spam = as.integer(factor(spam)) - 1))

## We need to do a littl more work to get predictions. This code adapted from
## http://stackoverflow.com/questions/11839886/r-predict-glm-equivalent-for-mcmcpackmcmclogit#40275910
predict_mcmc_logit <- function(model, newdata) {
    1 / (1 + exp(-model %*% do.call(rbind, c(1,newdata))))
}

pd2 <- predict_mcmc_logit(fit2, newdata = test[-1])

p2 <- apply(pd2, 2, mean)

confmat(test$spam,
        factor(p2 > .5,
               labels = c("email", "spam")))
```

Both the `rstanarm` and `MCMCpack` model fitting functions have options to change the number of burn-in iterations, the number of MCMC iterations and so forth. Refer to the package documentation for details.

Examining the simulations
-------------------------

It is a good idea to examine the `trace` and other diagnostic
information. For `MCMCpack` models the `plot` function will give you
basic diagnostics. For `rstanarm` (and other `rstan` models) the
`launch_shinystan` function launches an interactive dashboard that you
can use to display information about the model.


```{r}
## not run; only makes sense in interactive environment
## launch_shinystan(fit1)

plot(fit1)
```


```{r}
## not run; only makes sense in interactive environment
## launch_shinystan(fit1)
plot(fit2)
```

Your Turn: Predict Party Using Bayesian Logistic Regression
===========================================================


1. Fit a logistic regression model predicting political party from
   votes and generate a confusion matrix to assess the accuracy of the
   model.

2. Fit a Bayesian logistic regression model predicting political party from
   votes and generate a confusion matrix to assess the accuracy of the
   model. You may use either the `MCMCpack` package or the `rstanarm` package.

3. (Advance) Try to find a way of improving upon the classification
   accuracy obtained in 1 and 2 above.


General MCMC Programming Environments
=====================================

Finally we've reached level three. Here we specify models in terms of
likelihood functions and priors and let our chosen MCMC algorithm
search the parameter space. 

There are two main options for doing this in R: `R2OpenBUGS` and
`rstan`. I'm going to demonstrate `rstan` since I think it is where
the current momentum and developer interest currently is.


```{r}
library(rstan)

## specify model as a charactor vector
model <- '
data {
  int<lower=0>  N;
  vector[N] spam;
  vector[N] A1;
  vector[N] A2;
  vector[N] A3;
  vector[N] A4;
  vector[N] A5;
  vector[N] A6;
  vector[N] A7;
  vector[N] A8;
  vector[N] A9;
  vector[N] A10;
  vector[N] A11;
  vector[N] A12;
  vector[N] A13;
  vector[N] A14;
  vector[N] A15;
  vector[N] A16;
  vector[N] A17;
  vector[N] A18;
  vector[N] A19;
  vector[N] A20;
  vector[N] A21;
}
parameters {
  vector[22] beta;
  real<lower=0> sigma;
} 
model {
  spam ~ normal(beta[1] + beta[2] * A1 + beta[3] * A2 + beta[4] * A3
                    + beta[5] * A4 + beta[6] * A5 + beta[7] * A6 
                    + beta[8] * A7 + beta[9] * A8 
                    + beta[10] * A9 + beta[11] * A10
                    + beta[12] * A11 + beta[13] * A12
                    + beta[14] * A13 + beta[15] * A14
                    + beta[16] * A15 + beta[17] * A16
                    + beta[18] * A17 + beta[19] * A18
                    + beta[20] * A19 + beta[21] * A20
                    + beta[22] * A21,sigma);
}
'

## write model specification to file
cat(model, file = "spamModel.stan")


## prepair data (stan is picky...)
train$spam <- as.integer(train$spam) - 1
## stan doesn't like dots in variable names
names(train) <- gsub("\\.", "", names(train))
train.stan <- c(list(N = nrow(train)), as.list(train))

## fit the model
fit1b <- stan(file = "spamModel.stan",
              data = train.stan)


print(fit1)
```

```{r}

## plot the trace
traceplot(fit1b)

## plot parameters
plot(fit1b)
```

More information about Stan can be found at <http://mc-stan.org>.

